### 2018-01-10 submission/manuscript.Rmd
update discussion to include caveat on previously identified species

>>>>

One important caveat to this study is that even though specific genera associated with species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data, across study, to the genus level, it is not possible to clearly delineate what specific species are contributing to overall disease progression. However, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated micobe and our study clearly shows that regardless of whether the sample is tissue or stool, mouth-associated genera are commonly associated with an increased RR of both adenoma and carcinoma.

### 2018-01-10 submission/manuscript.tex
update discussion to include caveat on previously identified species

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}30,31{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis were included and only data sets that had sequences
available for analysis were included. Some studies were excluded because
they did not have publicly available sequences or did not have metadata
in which the authors were able to share and were excluded. After these
filtering steps, the following studies remained: Ahn, \emph{et al.}
{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}29{]},
Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
\emph{et al.} {[}23{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
al.} {[}22{]}, Hale, \emph{et al.} {[}26{]}, Kostic, \emph{et al.}
{[}32{]}, Lu, \emph{et al.} {[}25{]}, Sanapareddy, \emph{et al.}
{[}24{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}20{]},
and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}33{]} study was not
included because the 90 individuals analyzed within the study are
contained within the larger Baxter study {[}13{]}. After sequence
processing all the case samples for the Kostic study had 100 or less
sequences remaining and was not used which left a total of 14 studies
that analysis could be completed on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}12{]}) and directly from the authors (n = 4,
{[}18,20,24,26{]}). Each study was processed using the mothur (v1.39.3)
software program {[}34{]} and quality filtering utilized the default
methods, used in mothur, for either 454 or Illumina based sequencing. If
it was not possible to use the defaults, the stated quality cut-offs,
from the study itself, were used instead. Chimeras were identified and
removed using VSEARCH {[}35{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}36{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}37{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}38{]} and then Z-score
normalized using the car (v2.1.5) package {[}39{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}40{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hypervariable
region used. Relative risk was analyzed using both the epiR (v0.9.87)
and metafor (v2.0.0) packages {[}41,42{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.4) package {[}43{]}. Random Forest models were built using
both the caret (v6.0.77) and randomForest (v4.6.12) packages
{[}44,45{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}46,47{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}48,49{]}.

@@

\hypertarget{ref-keku_gastrointestinal_2015}{}
30. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
31. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
{[}cited 2017 Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-kostic_genomic_2012}{}
32. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-zackular_human_2014}{}
33. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
34. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}

\hypertarget{ref-rognes_vsearch_2016}{}
35. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
36. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
2017;2.

\hypertarget{ref-r_citation_2017}{}
37. R Core Team. R: A language and environment for statistical computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2017. Available from: \url{https://www.R-project.org/}

\hypertarget{ref-rcompanion_citation_2017}{}
38. Mangiafico S. Rcompanion: Functions to support extension education
program evaluation {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=rcompanion}

\hypertarget{ref-car_citation_2011}{}
39. Fox J, Weisberg S. An R companion to applied regression
{[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
\url{http://socserv.socsci.mcmaster.ca/jfox/Books/Companion}

\hypertarget{ref-lme4_citation_2015}{}
40. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
models using lme4. Journal of Statistical Software. 2015;67:1--48.

\hypertarget{ref-epir_citation_2017}{}
41. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
epidemiological data {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=epiR}

\hypertarget{ref-metafor_citation_2010}{}
42. Viechtbauer W. Conducting meta-analyses in R with the metafor
package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
Available from: \url{http://www.jstatsoft.org/v36/i03/}

\hypertarget{ref-vegan_citation_2017}{}
43. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
from: \url{https://CRAN.R-project.org/package=vegan}

\hypertarget{ref-caret_citation_2017}{}
44. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
Cooper T, et al. Caret: Classification and regression training
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=caret}

\hypertarget{ref-randomforest_citation_2002}{}
45. Liaw A, Wiener M. Classification and regression by randomForest. R
News {[}Internet{]}. 2002;2:18--22. Available from:
\url{http://CRAN.R-project.org/doc/Rnews/}

\hypertarget{ref-pwr_citation_2017}{}
46. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
2017. Available from: \url{https://CRAN.R-project.org/package=pwr}

\hypertarget{ref-statmod_citation_2016}{}
47. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
gaussian distribution. R Journal. 2016;8:339--51.

\hypertarget{ref-ggplot2_citation_2009}{}
48. Wickham H. Ggplot2: Elegant graphics for data analysis
{[}Internet{]}. Springer-Verlag New York; 2009. Available from:
\url{http://ggplot2.org}

\hypertarget{ref-gridextra_citation_2017}{}
49. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=gridExtra}

>>>>

One important caveat to this study is that even though specific genera
associated with species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean these specific species are not
important in human CRC {[}21,30{]}. Since we are limited in our
aggregation of the data, across study, to the genus level, it is not
possible to clearly delineate what specific species are contributing to
overall disease progression. However, \emph{Streptococcus gallolyticus}
subsp. \emph{gallolyticus} is a mouth-associated micobe and our study
clearly shows that regardless of whether the sample is tissue or stool,
mouth-associated genera are commonly associated with an increased RR of
both adenoma and carcinoma.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}31,32{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis were included and only data sets that had sequences
available for analysis were included. Some studies were excluded because
they did not have publicly available sequences or did not have metadata
in which the authors were able to share and were excluded. After these
filtering steps, the following studies remained: Ahn, \emph{et al.}
{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}29{]},
Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
\emph{et al.} {[}23{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
al.} {[}22{]}, Hale, \emph{et al.} {[}26{]}, Kostic, \emph{et al.}
{[}33{]}, Lu, \emph{et al.} {[}25{]}, Sanapareddy, \emph{et al.}
{[}24{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}20{]},
and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}34{]} study was not
included because the 90 individuals analyzed within the study are
contained within the larger Baxter study {[}13{]}. After sequence
processing all the case samples for the Kostic study had 100 or less
sequences remaining and was not used which left a total of 14 studies
that analysis could be completed on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}12{]}) and directly from the authors (n = 4,
{[}18,20,24,26{]}). Each study was processed using the mothur (v1.39.3)
software program {[}35{]} and quality filtering utilized the default
methods, used in mothur, for either 454 or Illumina based sequencing. If
it was not possible to use the defaults, the stated quality cut-offs,
from the study itself, were used instead. Chimeras were identified and
removed using VSEARCH {[}36{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}37{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}38{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}39{]} and then Z-score
normalized using the car (v2.1.5) package {[}40{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}41{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hypervariable
region used. Relative risk was analyzed using both the epiR (v0.9.87)
and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.4) package {[}44{]}. Random Forest models were built using
both the caret (v6.0.77) and randomForest (v4.6.12) packages
{[}45,46{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}47,48{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}49,50{]}.

@@

\hypertarget{ref-strep_Aymeric_2017}{}
30. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
et al. Colorectal cancer specific conditions promoteStreptococcus
gallolyticusgut colonization. Proceedings of the National Academy of
Sciences {[}Internet{]}. Proceedings of the National Academy of
Sciences; 2017;115:E283--91. Available from:
\url{https://doi.org/10.1073/pnas.1715112115}

\hypertarget{ref-keku_gastrointestinal_2015}{}
31. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
32. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
{[}cited 2017 Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-kostic_genomic_2012}{}
33. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-zackular_human_2014}{}
34. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
35. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}

\hypertarget{ref-rognes_vsearch_2016}{}
36. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
37. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
2017;2.

\hypertarget{ref-r_citation_2017}{}
38. R Core Team. R: A language and environment for statistical computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2017. Available from: \url{https://www.R-project.org/}

\hypertarget{ref-rcompanion_citation_2017}{}
39. Mangiafico S. Rcompanion: Functions to support extension education
program evaluation {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=rcompanion}

\hypertarget{ref-car_citation_2011}{}
40. Fox J, Weisberg S. An R companion to applied regression
{[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
\url{http://socserv.socsci.mcmaster.ca/jfox/Books/Companion}

\hypertarget{ref-lme4_citation_2015}{}
41. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
models using lme4. Journal of Statistical Software. 2015;67:1--48.

\hypertarget{ref-epir_citation_2017}{}
42. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
epidemiological data {[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=epiR}

\hypertarget{ref-metafor_citation_2010}{}
43. Viechtbauer W. Conducting meta-analyses in R with the metafor
package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
Available from: \url{http://www.jstatsoft.org/v36/i03/}

\hypertarget{ref-vegan_citation_2017}{}
44. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
from: \url{https://CRAN.R-project.org/package=vegan}

\hypertarget{ref-caret_citation_2017}{}
45. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
Cooper T, et al. Caret: Classification and regression training
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=caret}

\hypertarget{ref-randomforest_citation_2002}{}
46. Liaw A, Wiener M. Classification and regression by randomForest. R
News {[}Internet{]}. 2002;2:18--22. Available from:
\url{http://CRAN.R-project.org/doc/Rnews/}

\hypertarget{ref-pwr_citation_2017}{}
47. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
2017. Available from: \url{https://CRAN.R-project.org/package=pwr}

\hypertarget{ref-statmod_citation_2016}{}
48. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
gaussian distribution. R Journal. 2016;8:339--51.

\hypertarget{ref-ggplot2_citation_2009}{}
49. Wickham H. Ggplot2: Elegant graphics for data analysis
{[}Internet{]}. Springer-Verlag New York; 2009. Available from:
\url{http://ggplot2.org}

\hypertarget{ref-gridextra_citation_2017}{}
50. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
{[}Internet{]}. 2017. Available from:
\url{https://CRAN.R-project.org/package=gridExtra}
